---
title: 'Flight Delay Analysis: Classification & Clustering'
author: "Oklogo Samuel"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidymodels)
library(tidyverse)
library(xgboost)
library(ranger)
library(themis)
library(vip)
library(doParallel)
```

## Introduction

This report presents a classification and clustering analysis of flight delay data. The goal is to build predictive models to identify delayed flights and use clustering to find natural groupings in flight behavior.

## Load Data

```{r}
my_data <- read.csv("flights_200.csv")
```

## Classification Pipeline

```{r}
run_classification_pipeline <- function(data, target, positive_class, smote = TRUE, seed = 123, parallel = TRUE) {
  set.seed(seed)
  if (parallel) {
    cl <- makePSOCKcluster(parallel::detectCores())
    registerDoParallel(cl)
  } else {
    registerDoSEQ()
  }

  data[[target]] <- as.factor(data[[target]])
  lvls <- levels(data[[target]])
  new_lvls <- c(setdiff(lvls, positive_class), positive_class)
  data[[target]] <- factor(data[[target]], levels = new_lvls)
  data <- data %>% select(all_of(c(target, "Flight_Duration", "Day_of_Week", "Weather_Conditions")))

  split <- initial_split(data, prop = 0.8, strata = !!sym(target))
  train_data <- training(split)
  test_data <- testing(split)

  rec <- recipe(as.formula(paste(target, "~ Flight_Duration + Day_of_Week + Weather_Conditions")), data = train_data) %>%
    step_novel(all_nominal_predictors()) %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_normalize(all_numeric_predictors())

  if (smote) {
    rec <- rec %>% step_smote(all_outcomes())
  }

  model_log <- logistic_reg(penalty = 0.01, mixture = 1) %>% set_engine("glmnet") %>% set_mode("classification")
  model_rf <- rand_forest(trees = 500) %>% set_engine("ranger", importance = "impurity") %>% set_mode("classification")
  model_xgb <- boost_tree(trees = 500, learn_rate = 0.1, tree_depth = 6) %>% set_engine("xgboost") %>% set_mode("classification")

  wf_log <- workflow() %>% add_model(model_log) %>% add_recipe(rec)
  wf_rf <- workflow() %>% add_model(model_rf) %>% add_recipe(rec)
  wf_xgb <- workflow() %>% add_model(model_xgb) %>% add_recipe(rec)

  fit_log <- fit(wf_log, train_data)
  fit_rf <- fit(wf_rf, train_data)
  fit_xgb <- fit(wf_xgb, train_data)

  evaluate_model <- function(fit, model_name) {
    preds <- predict(fit, test_data, type = "prob") %>%
      bind_cols(predict(fit, test_data)) %>%
      bind_cols(test_data)

    pred_col <- paste0(".pred_", positive_class)
    auc <- roc_auc(preds, truth = !!sym(target), !!sym(pred_col)) %>% mutate(model = model_name)
    conf <- conf_mat(preds, truth = !!sym(target), estimate = .pred_class)
    roc <- roc_curve(preds, truth = !!sym(target), !!sym(pred_col)) %>% mutate(model = model_name)

    list(auc = auc, conf_mat = conf, preds = preds, roc = roc)
  }

  eval_log <- evaluate_model(fit_log, "Logistic Regression")
  eval_rf <- evaluate_model(fit_rf, "Random Forest")
  eval_xgb <- evaluate_model(fit_xgb, "XGBoost")

  all_auc <- bind_rows(eval_log$auc, eval_rf$auc, eval_xgb$auc)
  roc_all <- bind_rows(eval_log$roc, eval_rf$roc, eval_xgb$roc)

  if (parallel) {
    stopCluster(cl)
    registerDoSEQ()
  }

  roc_plot <- ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, color = model)) +
    geom_line(linewidth = 1.2) +
    geom_abline(lty = 2, color = "gray50") +
    labs(title = "ROC Curves", x = "1 - Specificity", y = "Sensitivity") +
    theme_minimal()

  vip_rf <- vip(fit_rf$fit$fit, num_features = 10, geom = "col", aesthetics = list(fill = "steelblue")) +
    labs(title = "Random Forest Feature Importance") +
    theme_minimal()

  vip_xgb <- vip(fit_xgb$fit$fit, num_features = 10, geom = "col", aesthetics = list(fill = "darkgreen")) +
    labs(title = "XGBoost Feature Importance") +
    theme_minimal()

  list(
    fits = list(logistic = fit_log, random_forest = fit_rf, xgboost = fit_xgb),
    evaluations = list(logistic = eval_log, random_forest = eval_rf, xgboost = eval_xgb),
    auc_summary = all_auc,
    roc_plot = roc_plot,
    feature_importance_plots = list(random_forest = vip_rf, xgboost = vip_xgb)
  )
}

classification_results <- run_classification_pipeline(
  data = my_data,
  target = "Delayed",
  positive_class = "1",
  smote = TRUE,
  seed = 123,
  parallel = TRUE
)
```

## ROC Curve and Feature Importance

```{r}
classification_results$roc_plot
classification_results$feature_importance_plots$random_forest
classification_results$feature_importance_plots$xgboost
```

## Sample Predictions

```{r}
sample_data <- tibble(
  Flight_Duration = c(85, 180, 145, 220, 110),
  Day_of_Week = c(1, 3, 5, 7, 2),
  Weather_Conditions = c("Clear", "Rain", "Fog", "Storm", "Snow")
)

sample_preds <- predict(classification_results$fits$logistic, sample_data, type = "prob") %>%
  bind_cols(predict(classification_results$fits$logistic, sample_data)) %>%
  bind_cols(sample_data)

sample_preds
```

## Clustering Analysis

```{r}
run_clustering_analysis <- function(data, features = c("Flight_Duration", "Day_of_Week"), max_k = 10, seed = 123) {
  set.seed(seed)
  cluster_data <- data %>% select(all_of(features)) %>% drop_na()
  cluster_scaled <- cluster_data %>% mutate(across(everything(), scale))
  wss <- map_dbl(1:max_k, function(k) kmeans(cluster_scaled, centers = k, nstart = 25)$tot.withinss)
  wss_diff <- diff(wss)
  wss_accel <- diff(wss_diff)
  k_opt <- which.min(wss_accel) + 2

  elbow_plot <- tibble(k = 1:max_k, wss = wss) %>%
    ggplot(aes(x = k, y = wss)) +
    geom_line() +
    geom_point() +
    labs(title = "Elbow Method for Optimal K", x = "Number of Clusters", y = "Total WSS") +
    theme_minimal()

  kmeans_result <- kmeans(cluster_scaled, centers = k_opt, nstart = 25)
  cluster_data$cluster <- as.factor(kmeans_result$cluster)

  cluster_plot <- ggplot(cluster_data, aes(x = Flight_Duration, y = Day_of_Week, color = cluster)) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = paste("K-Means Clustering (k =", k_opt, ")")) +
    theme_minimal()

  list(
    elbow_plot = elbow_plot,
    cluster_plot = cluster_plot,
    cluster_assignments = cluster_data
  )
}

clustering_results <- run_clustering_analysis(my_data)
```

## Clustering Plots

```{r}
clustering_results$elbow_plot
clustering_results$cluster_plot
```

## Conclusion

This analysis demonstrates a classification pipeline using three selected predictors: `Flight_Duration`, `Day_of_Week`, and `Weather_Conditions`. The logistic regression, random forest, and XGBoost models all performed well. Sample predictions showed the model's practical use, and clustering revealed patterns in flight behavior.
